---
title: "Somatic Variant Pipeline Report"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    page-layout: full
    css: custom.css
execute:
  echo: false
  warning: false
  message: false
params:
  sample_id: "Sample"
  metrics_dir: "metrics"
  plots_dir: "plots"
resources:
  - "plots/**"
  - "metrics/**"
---
```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(glue)
  library(knitr)
})

# Resolve paths robustly
resolve_dir <- function(p) {
  if (dir.exists(p)) return(normalizePath(p))
  candidate <- file.path(getwd(), p)
  if (dir.exists(candidate)) return(normalizePath(candidate))
  p
}

metrics_dir <- resolve_dir(params$metrics_dir)
plots_dir   <- resolve_dir(params$plots_dir)

available_plots <- if (dir.exists(plots_dir)) {
  list.files(plots_dir, pattern = "\\.(png|pdf)$", full.names = FALSE)
} else character(0)

# Flexible plot include by filename pattern
include_plot <- function(pattern, prefer = "png") {
  matches <- grep(pattern, available_plots, value = TRUE, ignore.case = TRUE)
  if (length(matches) == 0) {
    cat(glue("*Plot not found:* `{pattern}`  \n"))
    return(invisible(NULL))
  }
  if (prefer == "png") {
    pngs <- grep("\\.(png)$", matches, value = TRUE, ignore.case = TRUE)
    if (length(pngs) > 0) matches <- pngs
  }
  knitr::include_graphics(file.path(plots_dir, matches[1]))
}

read_tsv_safe <- function(path) {
  if (!file.exists(path) || file.info(path)$size == 0) return(NULL)
  tryCatch(readr::read_tsv(path, show_col_types = FALSE), error = function(e) NULL)
}

# Load metrics
counts_long   <- read_tsv_safe(file.path(metrics_dir, "variant_counts_by_stage.tsv"))
summary_stats <- read_tsv_safe(file.path(metrics_dir, "summary_stats.tsv"))
union_anno    <- read_tsv_safe(file.path(metrics_dir, "upset_matrix.tsv"))
bam_info    <- read_tsv_safe(file.path(metrics_dir, "vcf_dp_summary.tsv"))

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

# -----------------------------
# KPI computations (fallback-friendly)
# -----------------------------

# KPI 1: total unique variants across all callers (union)
kpi_total_union <- NA_integer_
if (!is.null(union_anno) && all(c("CHROM","POS","REF","ALT") %in% names(union_anno))) {
  kpi_total_union <- nrow(union_anno)
} else if (!is.null(summary_stats)) {
  hit <- summary_stats %>% filter(str_detect(metric, "union|k1|total_unique|total_variants"))
  if (nrow(hit) > 0) kpi_total_union <- suppressWarnings(as.integer(hit$value[1]))
}

# Extract values from the data frame (first row)
if (!is.null(bam_info) && nrow(bam_info) > 0) {
  dp_median <- bam_info$median_dp[1]
  dp_min <- bam_info$min_dp[1]
  dp_max <- bam_info$max_dp[1]
} else {
  dp_median <- "NA"
  dp_min <- "NA"
  dp_max <- "NA"
}


# KPI 2: callers + per-caller counts (choose representative stage for "before consensus")
caller_stage <- "norm"
caller_table <- NULL
if (!is.null(counts_long)) {
  stage_pref <- c("norm","split","raw")
  caller_stage <- (stage_pref[stage_pref %in% unique(counts_long$stage)][1]) %||% unique(counts_long$stage)[1]
  caller_table <- counts_long %>%
    filter(stage == caller_stage) %>%
    group_by(caller) %>%
    summarize(total = sum(count, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(total))
}

fmt_int <- function(x) {
  if (is.na(x)) return("NA")
  format(x, big.mark = ",", scientific = FALSE)
}
```

## Summary

<div class="kpi-grid">

<div class="kpi-card kpi-card-blue">
<h3>🔬 Basic Information</h3>
<div class="kpi-content">
<p><strong>Sample:</strong> `r params$sample_id`</p>
<p><strong>Median depth:</strong> `r dp_median`</p>
<p><strong>Minimum depth:</strong> `r dp_min`</p>
<p><strong>Maximum depth:</strong> `r dp_max`</p>
</div>
</div>

<div class="kpi-card kpi-card-green">
<h3>🧬 Variant Landscape</h3>
<div class="kpi-content">
<div class="kpi-number">`r fmt_int(kpi_total_union)`</div>
<p class="kpi-description">Total unique variants across all callers (CHROM–POS–REF–ALT union)</p>
</div>
</div>

<div class="kpi-card kpi-card-purple">
<h3>🧪 Caller Summary</h3>
<div class="kpi-content">
<p class="kpi-subtitle">Stage: <strong>`r caller_stage`</strong></p>
```{r callers-kpi-table}
#| echo: false
if (is.null(caller_table) || nrow(caller_table) == 0) {
  cat("*Caller table not available*")
} else {
  knitr::kable(
    caller_table %>% mutate(total = format(total, big.mark=",")),
    col.names = c("Caller", "Mutations"),
    align = c("l", "r")
  )
}
```
</div>
</div>

</div>

## Pipeline Overview

::: {.plot-row}

::: {.plot-container}

### Variant counts across pipeline stages
```{r fig1a}
#| out-width: "100%"
include_plot("variant_counts_by_stage")
```

:::

::: {.plot-container}

### Variant composition by caller
```{r fig1b-counts}
#| out-width: "100%"
include_plot("variant_counts_by_caller")
```

:::

:::

## Agreement and Overlap

::: {.plot-row}

::: {.plot-container}

### Upset plot of variants
```{r fig2a}
#| out-width: "100%"
include_plot("upset_callers")
```

:::

::: {.plot-container}

### Caller contribution
```{r fig2b}
#| out-width: "100%"
include_plot("caller_contribution")
```

:::

:::

## K-of-N behavior

::: {.plot-row}

::: {.plot-container}

### Upset plot of variants
```{r fig3a}
#| out-width: "100%"
include_plot("k_curve_combined")
```

:::

::: {.plot-container}

### Caller contribution
```{r fig3b}
#| out-width: "100%"
include_plot("k_curve_by_vartype_bar")
```

:::

:::

## Complex events

::: {.plot-row}

::: {.plot-container}

### Upset plot of variants
```{r fig4a}
#| out-width: "100%"
include_plot("mnv_vs_multiallelic_events")
```

:::

::: {.plot-container}

### Caller contribution
```{r fig4b}
#| out-width: "100%"
include_plot("event_status_pies_merged")
```

:::

:::